The Evolution of Phenotypically Invariant Genetic Networks

### 2018-11-09 ridge-review-responses.tex

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
Your manuscript is quite long already, and the reviewers' suggestions of
modifications and clarification, which need to be implemented, will make the
manuscript even longer. I however share the reviewers' opinion that the
manuscript's two parts are only loosely related. My suggestion is therefore to
publish them separately. Although connecting the systems biology and the
quantitative genetics parts is indeed an exciting endeavor, this is not really
achieved in the current version of the manuscript, and it would be more
profitable to first better describe each part separately. Regarding the
connexion between the two parts, please also pay particular attention to R2's
first specific comment about dimensionality.
\end{quote}

@@

\begin{point}{p11, 2nd paragraph}
    Try to better relate these different studies to yours
    (e.g., what you add, what they do and you do not). In addition to R1's
    suggestions, consider \citet{weinreich2013fishers} and \citet{blanquart2016epistasis}.
% Weinreich, D. M. and Knies, J. L. (2013), FISHER'S GEOMETRIC MODEL OF ADAPTATION MEETS THE FUNCTIONAL SYNTHESIS: DATA ON PAIRWISE EPISTASIS FOR FITNESS YIELDS INSIGHTS INTO THE SHAPE AND SIZE OF PHENOTYPE SPACE. Evolution, 67: 2957-2972. doi:10.1111/evo.12156
% François Blanquart and Thomas Bataillon Epistasis and the Structure of Fitness Landscapes: Are Experimental Fitness Landscapes Compatible with Fisher’s Geometric Model?  GENETICS Early online April 6, 2016
\end{point}

\reply{
}

\begin{point}{Hybridization:}
    What is $\mathcal{X}$?
\end{point}

\reply{
    A holdover from previous notation; it is $\allS$ (fixed). \revref
}

\begin{point}{p15}
    ``The importance of including neutral directions in these models, which
    is not usually done'' -- Do not lines of isofitness correspond to neutral
    directions, and aren't these neutral directions already included in those models?
\end{point}

\reply{
    Good point! We meant neutral directions \emph{within the optimal set},
    and have reworded this.
}

Yes, exactly. This is a natural model,
so we were surprised to not find an analytic treatment of it
in the literature.

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set $\allS$ is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

@@

Yes, exactly. This is a natural model,
so we were surprised to not find an analytic treatment of it
in the literature.

\begin{point}{}
    For system drift, you assume that ``Selection will tend to restrain this
motion, but movement along the optimal set $\allS$ is unconstrained, and so we expect
the population mean to drift along the optimal set like a particle diffusing.''
I see two major problems with this view. Both lead to slower divergence and
therefore run against your conclusions.
    \textit{(further points to follow)}
\end{point}

\reply{
}

\begin{point}{(drift along a ridge)}
    In the presence of epistasis, evolution on a neutral network (a high-fitness
    ridge) is \emph{not} due to drift alone, but also affected by (weak second order)
    selection in favor of mutational robustness / genetic canalization. In contrast
    to what you write, diffusion on the set of network coefficients corresponding
    to the optimal phenotype is not unbiased -- even if the optimal phenotypes do
    indeed all have the same fitness. Instead, selection drives the population to
    ``thicker'' parts of the network where the mean fitness of the population
    (including a cloud of mutants) is higher that on a narrow ridge. This is the
    basis of the evolution of robustness/canalization. It is possible to account
    for this effect, see Hermisson et al 2003. Amer. Nat. 161, 708-734;
    Alvarez-Castro et al, TPB 75 (2009) 109-122, or also Rice, 1998, Evolution 52,
    647-656; van Nimwegen, etal 1999, PNAS 96:9716–9720. Note that epistasis is
    necessary for the neutral evolution of incompatibilities, which is what you are
    aiming for.
\end{point}

@@

\reply{
}

\begin{point}{(drift along a ridge)}
    A second problem results from the fact that evolution on a high-fitness ridge
    often requires coordinated changes at many loci. Take the oscillator system
    that you use as an example: simultaneous changes at two genes are required to
    maintain the phenotype. This leads to a phenomenon called ``adaptive inertia''
    (see Baatz and Wagner, TPB 51, 49-66 and Alvarez-Castro etal . TPB 75, 109-122),
    which effectively slows down the movement along the ridge considerably. This
    problem applies, in particular, in a ``house-of-cards'' mutation regime when
    where rarely two mutations occur together on the same haplotype. In small
    populations, it typically requires that populations drift through shallow
    fitness valleys. While this is possible, it slows down the process.  
\end{point}

@@

\begin{point}{}
    Population isolates and genetic load: Isn't this exactly ``founder effect
speciation''?
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
In the second part of the paper, motivated by the previous results, the authors explore a general
quantitative genetics model in which populations can drift stochastically near a set of equivalent
and optimal systems. Since the optimal set (or evolutionary ridge) is not closed under averaging or
recombination, two isolated populations can drift apart and accumulate enough genetic differences
so that they do not produce any viable offspring. Using some heuristics, several expressions are
derived to quantify the accumulation of genetic incompatibilities.

@@

\reply{
}

\begin{point}{Fig 7.}
    I have one important issue with this figure (and the assumptions of the underly-
ing quantitative genetic model). If one wants to be consistent with the assumption that
parental populations drift along the evolutionary ridge, I think one would need to assume
that selection is strong enough to constraint the mean of the population on the surface. This
would presumably require that $\sigma S /\gamma \ll 1$. First, I think this assumption (or something
alternative to that) should be made explicit in the text. Secondly, it seems to me that this
assumption is not satisfied for panel A and C: under the range of parameters proposed by
the authors, the parental populations could easily drift away from the optimal set, and in
particular, the heuristics derived in the main text would not be satisfied.
\end{point}

\reply{
}

\begin{point}{}
    Finally, it would be worth mentioning several old and recent works relating genetic drift
to speciation : Yamagushi and Iwasa, several articles by Gavrilets et al. (I think several
citations are missing here), and Mirò Pina and Schertzer.
\end{point}

### 2018-11-09 ridge_evolution_paper.tex
started on responses
a bit more

% fitness as a fn of distance
\newcommand{\fit}{\mathcal{F}}
% fitness as a fn of A
\newcommand{\fitx}{\mathcal{F}}
% set of optimal coefficients
\newcommand{\optx}{\mathcal{X}}
% optimal phenotype
\newcommand{\optph}{\Phi_0}
% distance in phenotype space
\newcommand{\dph}{d}
% incompatibility
\newcommand{\Incompat}{\mathcal{I}}

@@

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses}  % name of the review responses file

@@

\begin{abstract}
Drift along a ridge.
\end{abstract}

@@

@@

We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_p)$,
and that the corresponding fitness is determined by the distance of its phenotype to optimum.
The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$.
In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
The fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
% this assumption does not entail a loss of generality since $D$ is arbitrary.
We assume that in the region of interest, the map $D$ is smooth
and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2017infinitesimal} --
the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
where $\varepsilon$ is a random Gaussian displacement due to random assortment of parental alleles.

%% REVIEW OF RELATED RESULTS
This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
Since we allow for substantial genetic variation within populations,
this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
which did not consider reproductive incompatibility.
Substantial work on speciation has been done under the assumption of a monomorphic population
whose trait is shifted by sequential fixation of alleles,
i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
This work has included both stationary optima (like we study)
and moving optima (i.e., adaptation) \citep[e.g.,][]{barton2001role,chevin2014niche}.
\citet{martin2014fishers} derived this model from a few general assumptions,
and used random matrix theory to calculate the distribution of fitness effects.
\citet{chevin2014niche} studied a general version with neutral directions,
and found that the rate of accumulation of reproductive isolation 
decreases with $N_e$ with a form that depends on trait dimension.
\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
could be explained by this model 
(although best if fitness took the form $\exp(-d^k)$
for some $k > 2$, where $d$ is distance to the optimum),
and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
In the context of a stationary optimum, 
the primary contribution to hybrid unfitness is segregation variance,
i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
due to drift having changed the genetic basis of the trait separately in the parental populations
(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
This turns out to be the main source of incompatibilities in the model we study as well,
even though in principle curvature of the optimal set also contributes
(as also seen in \citet{rosas2010cryptic}).
Our analysis relies on local approximations;
on longer time scales the appropriate model may share properties with
the ``holey'' landscapes of \citet{gavrilets2004fitness}.

@@

\paragraph{System drift}
% plr how fast does it drift
We work with a randomly mating population of effective size $N_e$. 
If the population variation has standard deviation $\sigma$ in a particular direction,
since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the standard deviation of the mean of a random sample \citep{lande1981models}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but movement along the optimal set $\allS$ is unconstrained,
and so we expect the population mean to drift along the optimal set like a particle diffusing.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depends on constraints imposed by selection and 
correlations between the genetic variation underlying different coefficients 
(the $G$ matrix \citep{arnold2008understanding}).
It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
a ``cloud'' of width $\sigma$ about the population mean, 
which moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.

@@

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{wright1968evolution,barton2001role,rosas2010cryptic}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

\paragraph{Fisher's geometric model}
Substantial analytical work on speciation has been done using Fisher's geometric model, 
which can also predict Haldane's rule and the regular increase of incompatibility with genetic distance
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
The model is similar to the quantitative genetics model we use 
to predict the rate of speciation, 
although work on Fisher's geometric model generally only models substitutions
between populations, neglecting within-species polymorphism.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue for the importance of including neutral directions in these models,
which is not usually done (but see \citet{chevin2014niche}).

>>>>

% fitness as a fn of distance
\newcommand{\fit}{\mathcal{F}}
% fitness as a fn of A
\newcommand{\fitx}{\mathcal{F}}
% optimal phenotype
\newcommand{\optph}{\Phi_0}
% distance in phenotype space
\newcommand{\dph}{d}
% incompatibility
\newcommand{\Incompat}{\mathcal{I}}

@@

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{ridge-review-responses}  % name of the review responses file

@@

\begin{abstract}
    It has been often argued that a common feature of biological evolution
    is that there are many distinct combinations of low-level traits
    that combine to produce the same high-level trait
    -- i.e., there is more than one way to do the same thing.
    If so, then the fitness landscape induced on the low-level traits has a ridge
    along which populations will, in principle,
    drift due to random demographic fluctuations.
    Here, we use one such model of a fitness ridge and the infinitesimal model
    to ask how fast populations are expected to drift,
    and find that this process is expected to reproducive incompatibilities
    between isolated populations over a number of generations proportional
    to the effective population size.
\end{abstract}

@@

Ridges in a fitness landscape
could be due to a variety of causes.
For instance, lower-level traits could be strictly redundant,
or could produce a higher-level trait additively.
There could be a trade-off between two traits under selection,
such as XXX
We were motivated to write this paper
by results from system theory
that explicitly describe a large parameter space
of linear systems with identical input--output relationships \citep{shiffman2018system}.

@@

We will suppose that each organism's phenotype is determined by its vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_p)$,
and that the corresponding fitness is determined by the distance of its phenotype to optimum.
The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$,
% In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
and the fitness of an organism with coefficients $x$ will be $\exp(-D(x)^2)$.
% this assumption does not entail a loss of generality since $D$ is arbitrary.
We assume that in the region of interest, the map $D$ is smooth
and that we can locally approximate $D(x)^2$ near the optimal set $\allS$ 
as a quadratic surface.
As above, an individual's coefficients are given by averaging its parentally inherited coefficients and adding random noise due to segregation and possibly new mutation.
Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton2017infinitesimal} --
the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
where $\varepsilon$ is a random Gaussian displacement due to random assortment of parental alleles.

%% REVIEW OF RELATED RESULTS
This fitness landscape is locally Gaussian with a rank-deficient covariance matrix.
Since we allow for substantial genetic variation within populations,
this model falls in the same class as \citet{lande1981models} and \citet{lande1983measurement},
which did not consider reproductive incompatibility.
Substantial work on speciation has been done under the assumption of a monomorphic population
whose trait is shifted by sequential fixation of alleles,
i.e., Fisher's ``geometric model'' \citep{fisher1930genetical,poon2000compensating}.
This work has included both stationary optima (like we study)
and moving optima (i.e., adaptation) \citep[e.g.,][]{barton2001role,chevin2014niche}.
\citet{martin2014fishers} derived this model from a few general assumptions,
and used random matrix theory to calculate the distribution of fitness effects.
\citet{chevin2014niche} studied a general version with neutral directions,
and found that the rate of accumulation of reproductive isolation 
decreases with $N_e$ with a form that depends on trait dimension.
\citet{fraisse2016genetics} showed that most recognized empirical patterns in the speciation literature
could be explained by this model 
(although best if fitness took the form $\exp(-d^k)$
for some $k > 2$, where $d$ is distance to the optimum),
and \citet{simon2017coadapted} further compared predictions from the model to empirical data.
In the context of a stationary optimum, 
the primary contribution to hybrid unfitness is segregation variance,
i.e., greater phenotypic variance in $F_2$ hybrids than in parentals
due to drift having changed the genetic basis of the trait separately in the parental populations
(\citet{chevin2014niche} refers to these as ``transgressive incompatibilities'').
This turns out to be the main source of incompatibilities in the model we study as well,
even though in principle curvature of the optimal set also contributes
(as also seen in \citet{rosas2010cryptic}).
Our analysis relies on local approximations;
on longer time scales the appropriate model may share properties with
the ``holey'' fitness landscapes of \citet{gavrilets2004fitness}.

@@

\paragraph{System drift}
% plr how fast does it drift
We work with a randomly mating population of effective size $N_e$. 
If the population variation has standard deviation $\sigma$ in a particular direction,
since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the standard deviation of the mean of a random sample \citep{lande1981models}.
Selection will restrain motion away from $\allS$,
but movement along the optimal set $\allS$ is relatively unconstrained,
and so as a first approximation, 
the population mean will drift along the optimal set like a particle diffusing.
The speed of this random walk
will be modified by two factors.
First, epistasis may require coordinated substitutions at several sites,
which may be nearly impossible if the individual substitutions are selected against strongly enough
in isolation.
This  XXX

% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depends on constraints imposed by selection and 
correlations between the genetic variation underlying different coefficients 
(the $G$ matrix \citep{arnold2008understanding}).
It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
a ``cloud'' of width $\sigma$ about the population mean, 
which moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.

@@

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    dotted lines give phenotypic distances to the optimum.
    A pair of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown in red,
    and the distribution of one type of $F_2$ is shown in purple
    (other types of $F_2$ are not shown; 
    some would be a similar distance to the other side of the optimal set).
    The distribution of $F_2$ hybrids is appropriate for mixed homozygotes
    if both traits have a simple, one-locus genetic basis,
    but there is variation within each population at that locus.
}
\end{figure}

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\allS$. \revpoint{AE}{2}
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
The distribution of $F_2$ hybrids will have mean at the average of the two populations,
but will have higher variance \citep{wright1968evolution,barton2001role,rosas2010cryptic}.
This ``segregation variance'' of $F_2$ hybrids 
can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown by \citet{slatkin1994segregation}
(also see Appendix \ref{apx:seg_cov} for a different derivation).
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$ hybrids will have variance $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
If the optimal set $\allS$ has dimension $q$,
using the polygenic model of appendix \ref{apx:seg_cov}, 
$\omega$ is proportional to the number of degrees of freedom: $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

\paragraph{Fisher's geometric model}
Substantial analytical work on speciation has been done using Fisher's geometric model, 
which can also predict Haldane's rule and the regular increase of incompatibility with genetic distance
\citep{barton2001role,fraisse2016genetics,simon2017coadapted}.
The model is similar to the quantitative genetics model we use 
to predict the rate of speciation, 
although work on Fisher's geometric model generally only models substitutions
between populations, neglecting within-species polymorphism.
Indeed, our model may provide an additional mechanistic context 
in which Fisher's geometric model is appropriate,
although the typical degree of pleiotropy
and stability of the $G$-matrix are unknown in practice.
Our observations argue that for questions related to divergence between populations 
it may be important to model a non-point optimum,
as is in \citet{chevin2014niche}. \revpoint{AE}{3}
